 

### 2021-05-17 vertical-thermal-review.Rmd
small edits, and knit in latest schematic

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
Across all **#** forested NEON sites, and in closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019] wind speeds increase with height. With daily maximum wind speeds averaging up to **#** $m s^-1^$ higher at the top of the NEON vertical profile than at the bottom

@@

**Vertical biophysical drivers reviewed above influence heat and water stress patterns, and leaf temperature $T_{leaf}$ distribution along the vertical gradient, which in turn shapes leaf metabolism (see following section).**
Along the biophysical dynamics vertically, leaves adjust to their environment by maintaining optimum $T_{leaf}$, for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. Therefore, leaves can be cooler than air at high temperature and warmer than air at low temperature (limited homeothermy) or can be drastically warmer than $T_{air}$ (megathermy). In some cases, $T_{leaf}$ can be equalled to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

### In-situ observations
**The principles outlined above, vertical microclimate profile (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and evapotranspiration (where is this discussed? and do we say anything about evaporation?) together shape in-situ $T_{leaf}$ observations (Fig. `r fig_leaf_T`).**

Particularly, in open-forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground can collectively increase $T_{leaf}$ [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$  relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** Closed-canopy forests, such as tropical and semi-deciduous tropical, and temperate forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016]. Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper-canopies highly susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]. 
Comparing forests, upper-canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper-canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].

### Leaf traits

@@

#### *(clarify intra- vs inter-specific variation - 1 short paragraph)*

>>>>

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
Across all **#** forested NEON sites, and in closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019] wind speeds increase with height. With daily maximum wind speeds averaging up to **#** m s^-1^ higher at the top of the NEON vertical profile than at the bottom

@@

**Vertical biophysical drivers reviewed above influence heat and water stress patterns, thereby influencing leaf temperature $T_{leaf}$ distributions along the vertical gradient.**
Along the biophysical gradients, leaves adjust to their environment by maintaining, when possible, a $T_{leaf}$ that is optimal for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be cooler than air at high temperature and warmer than air at low temperature (limited homeothermy) or can be drastically warmer than $T_{air}$ (megathermy). In some cases, $T_{leaf}$ can be equalled to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

### In-situ observations
**The principles outlined above, vertical microclimate profile (Fig. `r fig_NEON_vertical`) and leaf traits (see following section) together shape in-situ $T_{leaf}$ observations (Fig. `r fig_leaf_T`).** 
In open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground can collectively increase $T_{leaf}$ [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$  relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. 
Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** 
Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016]. 
Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$, making upper-canopies highly susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]. 
Comparing forests, upper canopies of temperate deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].

## Leaf traits

@@

### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
